---
title: "Oceanography and satellite data for SOG"
author: "Patrick Pata"
date: '2023-01-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load libraries and files
```{r}
library(here)
source(here("functions/R_package_check.R"))
prepareLibrary()
source(here("functions/QOL_toolkit.R"))

# Load net metadata 
load( here("data/net_metadata.RData") )

theme_set(theme_bw())
set_here(path = "..")

# PolySet data frames for mapping the Strait of Georgia from the PBSmapping package
data(nepacLL)
data(nepacLLhigh) #higher resolution maps
colnames(nepacLLhigh) <- c("group", "POS", "lon", "lat") # rename columns
colnames(nepacLL) <- c("group", "POS", "lon", "lat")

SOGmap <- ggplot() + 
  geom_polygon(data=nepacLLhigh, aes(lon, lat, group=group), 
               fill="grey85", size=0.2, color="black") +
  coord_map(projection='mercator', 
            xlim = c(-125.5, -122.5), ylim = c(48.5, 50.1)) + 
  theme(axis.title = element_blank(), axis.text = element_blank()) +
  theme_classic() 

```

# Subset Strait of Georgia satellite data
## subset data from matlab
```{r, eval = FALSE}
# # the SoG satellite data associated with net data
# load(here("data/SOG_satellite_acri_chl_2023_01_10.RData"))
# netsat <- satellite.chl.grid %>% 
#   distinct(lon, lat) %>% 
#   mutate(lonlat = paste0(lon,"_",lat))

# the complete satellite dataset generated from matlab
library(R.matlab)
fol <- "C:/Sync/UBC_codes/UBC-Zooplankton-Matlab-Codes/"
f <- "acri_chla_bc_long_table_19972022.mat"

data <- readMat(paste0(fol,f))
chl <- as.data.frame(data$acri.chl.sub)
colnames(chl) <- c("year","month","lon","lat","chl")

chl <- chl %>% 
  filter(lon >= -125.5 & lon <= -122.5 & lat >= 48.5 & lat <= 51) 

chl.grid <- chl %>% 
  distinct(lon, lat) %>% 
  mutate(keep = FALSE)

# find grid cells which are 0.12 degrees (3 grid cells) away from any of the net data points 
sog.net <- zoop.SOG.meta %>% 
  distinct(Longitude, Latitude) %>% 
  rename(lon = Longitude, lat = Latitude)
for (i in c(1:nrow(sog.net))) {
  ii <- which(chl.grid$lon-0.12 <= sog.net[i,]$lon &
              chl.grid$lon+0.12 >= sog.net[i,]$lon & 
              chl.grid$lat-0.12 <= sog.net[i,]$lat &
              chl.grid$lat+0.12 >= sog.net[i,]$lat)
  if(!is_empty(ii)) {
    chl.grid[ii,]$keep <- TRUE
  }
}

chl.grid <- chl.grid %>% 
  filter(keep == TRUE)

SOGmap +
  geom_point(data = chl.grid, aes(x = lon, y = lat))

satellite.sog <- chl.grid %>% 
  filter(keep == TRUE) %>% 
  select(-keep) %>% 
  left_join(chl, by = c("lon","lat")) %>% 
  arrange(month) %>% 
  arrange(year)

# Export the subset to RData file
save(satellite.sog, file = here("data/SOG_satellite_chl_19972022.RData"))
```

# -------------- old codes -----------------------
```{r}
# Subset grid cells by matching grid cells with net data
chl.sog <- chl %>% 
  mutate(lon2 = as.numeric(format(round(lon, 2), nsmall = 2)),
         lat2 = as.numeric(format(round(lat, 3), nsmall = 3)))
  mutate(lonlat = paste0(lon2,"_",lat2)) %>%
  filter(lonlat %in% netsat$lonlat)

# # Subset grid cells by matching lon and lat points in net data
# chl.sog <- chl %>% 
#   mutate(lon2 = as.numeric(format(round(lon, 2), nsmall = 2)),
#          lat2 = as.numeric(format(round(lat, 3), nsmall = 3))) %>%
#   filter(lon2 %in% netsat$lon,
#          lat2 %in% netsat$lat)

AA <- chl.sog %>% 
  distinct(lon,lat, .keep_all = TRUE) 
# with missing grid cells
BB <- AA


g <- SOGmap +
  geom_point(data = distinct(AA, lon2, lat2), 
             aes(x = lon2, y = lat2))
g

# library(plotly)
# ggplotly(g)

# save(satellite.chlorophyll,
#      file = here("data/SOG_satellite_chl_19972022.RData"))

# rm(netsat)
```


```{r}
group_by(year, month) %>% 
  summarise(chl.mean = mean(chl, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(YearMonth = year + (month-1)/12) %>% 
  # calculate anomaly
  group_by(month) %>% 
  mutate(month.mean = mean(chl.mean),
         chl.anom = chl.mean - month.mean) %>% 
  ungroup()

g1 <- ggplot(chl.sog, aes(x = YearMonth, y = chl.mean)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm") +
  scale_y_continuous(trans = "log") +
  ggtitle("Offshore GOA - Station P") + xlab("")

g2 <- ggplot(chl.sog, aes(x = YearMonth, y = chl.anom, fill = sign)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(1880,2020,5)) +
  scale_color_manual(values = c("firebrick","cyan4")) +
  xlab("Year") +
  theme(legend.position = "none") 
  # ggtitle("Offshore GOA - Station P")
g1
g2


# Save to csv file
chl.stnP.csv <- chl.stnP %>% 
  select(YearMonth, Chlorophyll = chl.mean)
write.csv(chl.stnP.csv, file = "output/chlorophyll_station_p.csv",
          row.names = FALSE)

png("figures/chlorophyll_station_p.png", width = 6, height = 3, 
    units = "in",res = 300)
ggplot(chl.stnP.csv, aes(x = YearMonth, y = Chlorophyll)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(1880,2020,20)) +
  ylab("Chlorophyll-a Station P") + xlab("Year") 
dev.off()

```





# Satellite data
```{r}
load(here("data/SOG_satellite_chl_19972022.RData"))

# View which grid cells have net data
sat.grid <- satellite.chlorophyll %>% 
  distinct(lon, lat)
SOGmap +
  geom_point(data = distinct(sat.grid, lon, lat), 
             aes(x = lon, y = lat))
```


# Calculating the average chlorophyll time series from gridded satellite data
Chlorophyll a and other ocean color variables could be collected from satellites. Here, we provide gridded satellite-derived chlorophyll values for the Strait of Georgia.
```{r}
colnames(satellite.chl.grid)

SOGmap +
  geom_point(data = distinct(satellite.chl.grid, lon, lat), 
             aes(x = lon, y = lat))

chl.timeseries <- satellite.chl.grid %>% 
  # The coverage of grid cells with chlorophyll data varies every month due to cloud cover. For calculating the SOG chlorophyll average, only include grid cells which are often sampled throughout the time series, set to an arbitrary threshold of N data points = 100.
  group_by(lon, lat) %>% 
  mutate(Ndata = sum(!is.na(chl))) %>% 
  ungroup() %>% 
  filter(Ndata >= 100) %>% 
  # Calculating the average chlorophyll for SOG. Also divide into northern and central SOG roughly at 49.5 degrees latitude.
  # overall average
  group_by(year, month) %>% 
  mutate(chla.sog = mean(chl, na.rm = T),
         chla.sog.sd = sd(chl, na.rm = T)) %>% 
  # regional average
  mutate(region = if_else(lat >= 49.5, "northern", "central")) %>% 
  group_by(year, month, region) %>% 
  mutate(chla.sog.region = mean(chl, na.rm = T),
         chla.sog.region.sd = sd(chl, na.rm = T)) %>% 
  ungroup() %>% 
  # summarise
  select(-c(lon,lat,chl,Ndata)) %>% 
  distinct() %>% 
  # create year.month variable for timeseries
  mutate(year.month = year + (month-1)/12)

# Plot the timeseries
# overall
ggplot(data = distinct(chl.timeseries, year.month, .keep_all = T ),
       aes(x = year.month, y = chla.sog)) +
  geom_point() +
  geom_line() 
  # scale_y_continuous(trans='log10') 

# per region
ggplot(data = chl.timeseries,
       aes(x = year.month, y = chla.sog.region,
           color = region)) +
  geom_point() 
  

```

# Anomally time series
For time series data, it is often interesting to look into the anomaly of a parameter relative to a climatological mean. Here we calculate the monthly climatology of chla in the SOG and then, calculate the anomaly time series.
```{r}
# Calculate the monthly climatology of the entire SOG
chl.timeseries <- chl.timeseries %>% 
  group_by(month) %>% 
  mutate(chla.sog.clim = mean(chla.sog, na.rm = T)) %>% 
  ungroup() %>% 
  # Calculate the anomaly time series
  mutate(chla.sog.anom = chla.sog - chla.sog.clim)

# Plot anomaly as bar plots relative to zero
ggplot(data = distinct(chl.timeseries, year.month, .keep_all = T ),
       aes(x = year.month, y = chla.sog.anom)) +
  geom_bar(stat = "identity")
```

Open data points of dataset
```{r}
load( here("data/SOG_satellite_acri_chl_2023_01_10.RData"))
satellite.chl.grid

chl.sog <- chl %>% 
  filter(lon >= -125.5 & lon <= -122.5 & lat >= 48.5 & lat <= 51) 

# grids with net data
A <- satellite.chl.grid %>% 
  distinct(lon, lat) %>% 
  mutate(lonlat = paste0(lon,"_",lat)) %>% 
  mutate(orig = T)

chl.sog.2 <- chl.sog %>% 
  mutate(lon2 = as.numeric(format(round(lon, 2), nsmall = 2)),
         lat2 = as.numeric(format(round(lat, 3), nsmall = 3))) %>%
  mutate(lonlat = paste0(lon2,"_",lat2)) %>% 
  filter(lonlat %in% A$lonlat)


SOGmap +
  geom_point(data = distinct(chl.sog.2, lon, lat), 
             aes(x = lon, y = lat))
```








