---
title: "Exploring the zooplankton dataset file"
author: "Patrick Pata"
date: '2023-01-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and files

```{r}
library(here)
source(here("functions/R_package_check.R"))
prepareLibrary()
source(here("functions/toolkit.R"))

theme_set(theme_bw())
set_here(path = "..")

# Taxonomy list for matching with trait dataset
taxonomy <- openxlsx::read.xlsx(here("tables/taxonomy_table_20230207.xlsx"))

# PolySet data frames for mapping the Strait of Georgia from the PBSmapping package
data(nepacLL)
data(nepacLLhigh) #higher resolution maps
colnames(nepacLLhigh) <- c("group", "POS", "lon", "lat") # rename columns
colnames(nepacLL) <- c("group", "POS", "lon", "lat")

```

# Load and wrangle the zooplankton net data.

Data source for zooplankton abundance and biomass data: <https://open.canada.ca/data/en/dataset/2822c11d-6b6c-437e-ad65-85f584522adc> Data dictionary: <https://pacgis01.dfo-mpo.gc.ca/FGPPublic/Zooplankton_Data_SoG/Data_dictionary_EN_FR_SoG_zooplankton_data_1996-2018.htm>

Load the zooplankton abundance dataset from net observations

```{r}

# TODO: reorder factor levels for traitID and traitName
zoop.SOG <- read.csv(here("data/Zoopl_SofG_1996-2018_EN.csv"),
                            check.names = F) %>% 
  # Rename some columns
  rename("Longitude" = "Longitude(\xb0N)", "Latitude" = "Latitude(\xb0W)",
         "Mesh_size(um)" = "Mesh_size(\xb5m)",
         "Abundance(indiv/m3)" = "Abundance(m3)") %>% 
  
  # Create new scientific name column.
  mutate(scientificName = paste(Genus,Species)) %>% 
  mutate(scientificName = cleanScientificName(scientificName)) %>% 
  relocate(scientificName) %>% 
  
  # Calculate biomass as number of individuals per sample * average biomass per individual
  mutate(`Biomass(mg/m3)` = `Abundance(indiv/m3)` * `Average_biomass_per_individual(mg)`) %>% 
  
  # Assign seasons: The spring bloom occurs around March in the Strait of Georgia and so "spring" is set to be March-April-May
  mutate(Season = "Winter") %>% 
  mutate(Season = if_else(Month >= 3 & Month <=5, "Spring", Season),
         Season = if_else(Month >= 6 & Month <=8, "Summer", Season),
         Season = if_else(Month >= 9 & Month <=11, "Fall", Season)) %>% 
  
  # Assign date-time using lubridate
  mutate(datetime = lubridate::make_datetime(Year, Month, Day)) %>% 
  mutate(dayofyear = lubridate::yday(datetime)) %>% 
  
  # Need to include time of day as well     
                                             # as.numeric(unlist(strsplit(Time,split=":"))[1]),
                                             # as.numeric(unlist(strsplit(Time,split=":"))[2])))
  #  tz = "Canada/Pacific")) # setting the time zone now returns an error. Need to figure out why.
  
  # Exclude microzooplankton
  filter(Phylum %notin% c("Pyrrophycophyta (Division)","Ciliophora",
                            "Alveolates","Unclassified","Protozoa")) %>% 
  
  # Remove a zero at the 8th character to match with the old index names
  mutate(NetKey = paste0(str_sub(Index,1,7),str_sub(Index,9,)))

head(zoop.SOG)
```

# Explore data contents

The taxonomic resolution of the dataset is at species-stage or size class. See the data dictionary for details. The plankton data was provided as abundance (number of individuals per volume of water) and using the provided value of the average biomass per individual (specific to a species and stage or size class), the biomass is calculated.

View data variables: (For details, see: <https://pacgis01.dfo-mpo.gc.ca/FGPPublic/Zooplankton_Data_SoG/Data_dictionary_EN_FR_SoG_zooplankton_data_1996-2018.htm>)

```{r}
colnames(zoop.SOG)
```

View species present in the daataset:

```{r}
zoop.SOG$scientificName %>%
  unique() %>%
  tibble()
```

Create a species-level summary of the zooplankton dataset. This removes life stage or size class information so better to curate which life stages to include in an analysis before calculating species averages:

```{r}
zoop.SOG.sp <- zoop.SOG %>% 
  select(-c("Average_biomass_per_individual(mg)","Size_class_or_life_stage")) %>% 
  group_by(scientificName) %>% 
  mutate(`Abundance(indiv/m3)` = sum(`Abundance(indiv/m3)`),
         `Biomass(mg/m3)` = sum(`Biomass(mg/m3)`)) %>% 
  ungroup() %>% 
  distinct()

zoop.SOG.sp
# TODO: extract a column of life stages
```

Create and save a metadata file for the zooplankton data:

```{r}
zoop.SOG.meta <- zoop.SOG %>% 
  select(Index, Region, Station, Longitude, Latitude,
         Year, Month, Day, Time, Season, datetime, dayofyear,
         Twilight, Net_type,
         `Net_Diameter(m)`, `Mesh_size(um)`, 
         `Tow_start_depth(m)`,`Tow_end_depth(m)`,
         `Bottom_depth(m)`,`Volume(m3)`,
         NetKey) %>% 
  distinct()


# Output zooplankton metadata file
save(zoop.SOG.meta, file = here("data/net_metadata.RData"))

# Number of samples
length(unique(zoop.SOG.meta$NetKey))
```

There are available data from 1996 to 2018 but seasonal sampling frequencies vary per year:

```{r}
sample.summary <- zoop.SOG.meta %>% 
  group_by(Year, Season) %>% 
  count()
kable(sample.summary)
```

Which stations get sampled the most?

```{r}
zoop.SOG.meta %>% 
  group_by(Station) %>% 
  count() %>% 
  arrange(-n)
```

# Visualize the distribution of samples in space and time

Plot map of Strait of Georgia:

```{r}
# This is the high resolution base map for the SoG. Please change data = "nepacLL" if a lower resolution without some small islands is preferred.
SOGmap <- ggplot() + 
  geom_polygon(data=nepacLLhigh, aes(lon, lat, group=group), 
               fill="grey85", size=0.2, color="black") +
  coord_map(projection='mercator', 
            xlim = c(-126, -122), ylim = c(48, 51)) + 
  theme(axis.title = element_blank(), axis.text = element_blank()) +
  theme_classic()
SOGmap
```

Add sampled stations to the map:

```{r}
# Spatial distribution of samples. Note that there are multiple overlapping points and there are a few stations that are regularly sampled.
SOGmap +
  geom_point(data = zoop.SOG.meta, aes(x = Longitude, y = Latitude,
                                       color = `Bottom_depth(m)`)) +
  ggtitle("Distribution of zooplankton net samples")
```

Plot the temporal distribution of samples:

```{r}
ggplot(data = zoop.SOG.meta, aes(x = dayofyear, y = Year)) +
  geom_point(aes(color = Twilight)) +
  # Roughly mark seasons
  geom_vline(xintercept = c(32,120,212,304), linetype="dotted", color = "blue")
```

# Sample-level exploration

```{r}
# Calculate total abundance, biomass, and diversity metrics for each sample
zoop.SOG.summary <- zoop.SOG.sp %>% 
  group_by(Index) %>% 
  summarise(Richness = specnumber(`Abundance(indiv/m3)`),
            Diversity.Shannon = diversity(`Abundance(indiv/m3)`, index = "shannon"),
            Diversity.Simpson = diversity(`Abundance(indiv/m3)`, index = "simpson")) %>% 
  mutate(Evenness = Diversity.Shannon/log(Richness))

zoop.SOG.summary
```
