---
title: "Oceanography and satellite data for SOG"
author: "Patrick Pata"
date: '2023-01-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load libraries and files
```{r}
library(here)
source(here("functions/R_package_check.R"))
prepareLibrary()
source(here("functions/QOL_toolkit.R"))

theme_set(theme_bw())
set_here(path = "..")

# Taxonomy list for matching with trait dataset
taxonomy <- openxlsx::read.xlsx(here("table_inputs/taxonomy_table_20221122.xlsx"))

# Load net metadata 
load( here("data_outputs/net_metadata.RData") )

# Load SOG physics data
load( here("data_inputs/SOG_physics_data_2023_01_10.RData"))

# Load gridded satellite CHL data
load( here("data_inputs/SOG_satellite_acri_chl_2023_01_10.RData"))

# PolySet data frames for mapping the Strait of Georgia from the PBSmapping package
data(nepacLL)
data(nepacLLhigh) #higher resolution maps
colnames(nepacLLhigh) <- c("group", "POS", "lon", "lat") # rename columns
colnames(nepacLL) <- c("group", "POS", "lon", "lat")
```

Filter the physical data
```{r}
physics.ctd <- physics.ctd %>% 
  dplyr::select(NetKey, CTD, CTDKey, BottleKey, ctdnum, 
                TempI10, TempI50, TempINet, 
                SaliI10, SaliI50, SaliINet,
                DensI10, DensI50, DensINet,
                OxygI10, OxygI50, OxygINet)
```


# Distribution of CTD samples
A common sensor in oceanographic studies is the CTD (conductivity, temperature, depth). Conductivity is used to calculate for salinity. Additional sensors, such as a dissolved oxygen sensor, are often attached to the CTD system. A CTD sensor is typically deployed from research vessels and drops to a target depth, often just above the bottom of the ocean. It logs physical variables along the way and these are generally interpolated into 1 m depth bins. 

The physics dataset includes CTD data for salinity, temperature, and oxygen. Density was calculated from the temperature, salinity, and pressure information. The data are provided as depth-integrated average values from the surface to either 10 m, 50 m, or net tow depth. The variables have a suffix of I10, I50, or INet corresponding to the integration depth.

Here, we will first explore the spatial and temporal distribution of the CTD data. Notice that the CTD data were collected all over the Strait of Georgia and there are may overlapping points. Both spatially and temporally, the sampling frequency is uneven. Although there are permanent stations which are regularly monitored.

```{r}
# Structure of the physics data
colnames(physics.ctd)

# Subset columns with CTD data
ctd.SOG <- physics.meta %>% 
  filter(!is.na(CTDKey))
  
# This is the high resolution base map for the SoG. Please change data = "nepacLL" if a lower resolution without some small islands is preferred.
SOGmap <- ggplot() + 
  geom_polygon(data=nepacLLhigh, aes(lon, lat, group=group), 
               fill="grey85", size=0.2, color="black") +
  coord_map(projection='mercator', 
            xlim = c(-126, -122), ylim = c(48, 51)) + 
  theme(axis.title = element_blank(), axis.text = element_blank()) +
  theme_classic() 

# Spatial distribution of samples. Note that there are multiple overlapping points and there are a few stations that are regularly sampled.
SOGmap +
  geom_point(data = ctd.SOG, aes(x = Longitude, y = Latitude,
                                       color = `Depth_Bottom.m.`),
             alpha = 0.5) +
  ggtitle("Distribution of CTD samples")

# Temporal distribution of samples
ggplot(data = ctd.SOG, aes(x = dayofyear, y = Year)) +
  geom_point(aes(color = Twilight), alpha = 0.5) +
  # Roughly mark seasons
  geom_vline(xintercept = c(32,120,212,304), linetype="dotted", color = "blue")

# List most common research stations
ctd.SOG %>% 
  group_by(Station) %>% 
  count() %>% 
  arrange(-n) 
  head(10)
  
# Update the map and temporal distribution to reflect these stations
# Create a new variable Station2 which indicates if in the three main stations of not
ctd.SOG <- ctd.SOG %>% 
  mutate(Station2 = if_else(Station %in% c("GEO1","CPF1","CPF2"),
                            Station, "Other"))
SOGmap +
  geom_point(data = ctd.SOG, aes(x = Longitude, y = Latitude,
                                       color = Station2),
             alpha = 0.5) +
  ggtitle("Distribution of CTD samples")

ggplot(data = ctd.SOG, aes(x = dayofyear, y = Year)) +
  geom_point(aes(color = Station2), alpha = 0.5, size = 2) +
  # Roughly mark seasons
  geom_vline(xintercept = c(32,120,212,304), linetype="dotted", color = "blue")

```


# Time series of ctd data
Here, we do a simple curation of the physics data and present a monthly averaged CTD time series.
```{r}
ctd.ts <- physics.ctd %>% 
  # Merge some metadata
  left_join(distinct(ctd.SOG, CTDKey, Station, Longitude, Latitude,
                     Year, Month, Day), by = "CTDKey") %>% 
  # Here, we will only consider plotting the time series of the 3 most common stations.
  filter(Station %in% c("GEO1","CPF1","CPF2")) %>% 
  # Lengthen the data frame
  pivot_longer(cols = c("TempI10", "TempI50", "TempINet", 
                        "SaliI10", "SaliI50", "SaliINet",
                        "DensI10", "DensI50", "DensINet",
                        "OxygI10", "OxygI50", "OxygINet"),
               names_to = "Variable", values_to = "Value") %>% 
  filter(!is.na(Value)) %>% 
  # Calculate monthly averages
  group_by(Year, Month, Variable) %>% 
  summarise(Value = mean(Value), n = n()) %>% 
  ungroup() %>% 
  # Derive Year-Month variable for plotting
  mutate(YearMonth = Year + (Month - 1)/12)

# Visualize the time series for the upper 50 m  CTD values
ggplot(data = filter(ctd.ts, Variable %in% c("TempI50","SaliI50","DensI50","OxygI50")),
       aes(YearMonth, Value)) + 
  geom_point() +
  facet_wrap(~Variable, ncol = 1, scales = "free_y")

```



# Calculating the average chlorophyll time series from gridded satellite data
Chlorophyll a and other ocean color variables could be collected from satellites that orbit the Earth. Here, we provide gridded satellite-derived chlorophyll values for the Strait of Georgia. 
```{r}
colnames(satellite.chl.grid)

SOGmap +
  geom_point(data = distinct(satellite.chl.grid, lon, lat), 
             aes(x = lon, y = lat))

chl.timeseries <- satellite.chl.grid %>% 
  # The coverage of grid cells with chlorophyll data varies every month due to cloud cover. For calculating the SOG chlorophyll average, only include grid cells which are often sampled throughout the time series, set to an arbitrary threshold of N data points = 100.
  group_by(lon, lat) %>% 
  mutate(Ndata = sum(!is.na(chl))) %>% 
  ungroup() %>% 
  filter(Ndata >= 100) %>% 
  # Calculating the average chlorophyll for SOG. Also divide into northern and central SOG roughly at 49.5 degrees latitude.
  # overall average
  group_by(year, month) %>% 
  mutate(chla.sog = mean(chl, na.rm = T),
         chla.sog.sd = sd(chl, na.rm = T)) %>% 
  # regional average
  mutate(region = if_else(lat >= 49.5, "northern", "central")) %>% 
  group_by(year, month, region) %>% 
  mutate(chla.sog.region = mean(chl, na.rm = T),
         chla.sog.region.sd = sd(chl, na.rm = T)) %>% 
  ungroup() %>% 
  # summarise
  select(-c(lon,lat,chl,Ndata)) %>% 
  distinct() %>% 
  # create year.month variable for timeseries
  mutate(year.month = year + (month-1)/12)

# Plot the timeseries
# overall
ggplot(data = distinct(chl.timeseries, year.month, .keep_all = T ),
       aes(x = year.month, y = chla.sog)) +
  geom_point() +
  geom_line() 
  # scale_y_continuous(trans='log10') 

# per region
ggplot(data = chl.timeseries,
       aes(x = year.month, y = chla.sog.region,
           color = region)) +
  geom_point() 
  

```

# Anomally time series
For time series data, it is often interesting to look into the anomaly of a parameter relative to a climatological mean. Here we calculate the monthly climatology of chla in the SOG and then, calculate the anomaly time series.
```{r}
# Calculate the monthly climatology of the entire SOG
chl.timeseries <- chl.timeseries %>% 
  group_by(month) %>% 
  mutate(chla.sog.clim = mean(chla.sog, na.rm = T)) %>% 
  ungroup() %>% 
  # Calculate the anomaly time series
  mutate(chla.sog.anom = chla.sog - chla.sog.clim)

# Plot anomaly as bar plots relative to zero
ggplot(data = distinct(chl.timeseries, year.month, .keep_all = T ),
       aes(x = year.month, y = chla.sog.anom)) +
  geom_bar(stat = "identity")
```


